---
title: "tut4"
author: "Tom Lyle"
output: html_document
---

## Question 1
Given the following vectors;

```{r, include=TRUE, eval=TRUE}
vecX <- c(3, 9, -1, 4, 2, 6)
vecY <- c(5, 2, 0, 9, 3, 4)
```
create a chunk of code to perform these operations:

1. Create a matrix matX, from the vector (vecX), where number of rows are 2
```{r, include=TRUE, eval=TRUE}
matX <- matrix(vecX, nrow = 2)
```
2. Create a matrix matY, from the vector (vecY), where number of rows are 2,
```{r, include=TRUE, eval=TRUE}
matY <- matrix(vecY, nrow = 2)
```
3. Display the values of elements of the first row from matrix matX.
```{r, include=TRUE, eval=TRUE}
matX[1,]
```
4. Display the value of the element in the second row and third column of matY,
```{r, include=TRUE, eval=TRUE}
matY[2,3]
```
5. Sum up the two matrices and display the output,
```{r, include=TRUE, eval=TRUE}
matX + matY
```
6. Subtract the matY from matX and display the output,
```{r, include=TRUE, eval=TRUE}
matX - matY
```
7. Multiply the two matrices (element-wise) and print the output,
```{r, include=TRUE, eval=TRUE}
matX * matY
```
8. Transpose the matY and assign the output to a third matrix called matZ,
```{r, include=TRUE, eval=TRUE}
matZ <- t(matY)
```
9. Perform matrix multiplication between matX and matZ and assign the output to matW,
```{r, include=TRUE, eval=TRUE}
matW <- matX %*% matZ
```
10. Display the dimension of the matW matrix,
```{r, include=TRUE, eval=TRUE}
dim(matW)
```
11. Display the maximum, minimum and average value of matrix, matW
```{r, include=TRUE, eval=TRUE}
max(matW)
min(matW)
mean(matW)
```
12. Reflect and discuss with your colleagues what matrices are and when we should use them.

2-dimensional structures, used well for numbers, photos, chess boards and other 2-dimensional representations.

## Data frames
```{r, eval=TRUE, include=TRUE}
authors=c("Reda", "John", "Edward", "Ben")
year=c(2104, 2016, 2005, 2019)
publishers=c("Wiley", "Springer", "Sage", "International Books")

books <- data.frame(authors=authors, year= year, publishers= publishers)

# print the data frame entries
print(books)

# print the values of the authors entry
books$authors
# or ...
books["authors"] # what will happen if you use double square brackets? books[["authors"]] 
books[["authors"]] # Back to vector!
```

## Question 2

1. Create your own data frame that represents a tiny database with one table holding information about 5 movies. The variables (i.e., columns) of this table include the movies’ title, director, year, and country.
```{r, eval=TRUE, include=TRUE}
movie.title = c("Interstellar", "How to Train Your Dragon", "Baby Driver", "Spiderman: Into the Spiderverse", "Spirited Away")
movie.director = c("Christopher Nolan", "Chris Sanders", "Edgar Wright", "Bob Persichetti", "Hayao Miyazaki")
movie.year = c(2014, 2010, 2017, 2018, 2001)
movie.country = c("U.S.A", "Berk", "U.S.A", "Marvel's U.S.A", "Japan")

filmsA <- data.frame(title=movie.title,
                     director=movie.director,
                     year=movie.year,
                     country=movie.country)
```

2. Create another data frame of the same format as above but containing just one new movie.
```{r, eval=TRUE, include=TRUE}

filmsB <- data.frame(title=c("Whiplash"),
                     director=c("Damien Chazelle"),
                     year=c(2014),
                     country=c("U.S.A"))

```

3. Merge the two data frames row-wise using rbind() function.
```{r, include=TRUE, eval=TRUE}
filmsComb <- rbind(filmsA, filmsB)
```
4. Extract the titles of the merged data frame using the access operator, $
```{r, eval=TRUE, include=TRUE}
filmsComb$title
```
5. Try sorting the year using sort() function
```{r, eval=TRUE, include=TRUE}
filmsComb[sort(filmsComb$year),]
```
6. Replace the sort with the order() function, what is the difference between using sort and order?
```{r, eval=TRUE, include=TRUE}
filmsComb[order(filmsComb$year),]
```

